Comparison Of Several Sequential And Parallel Derivative-free Global Optimization Algorithms

### 2019-03-31 2019/PAVT/slides/slides.tex
PAVT: update slides

\begin{frame}
  \begin{center}

  \frametitle{Dimension reduction}
  Peano-type curve \(y(x)\) allows to reduce the dimension of the original problem:
  \begin{gather}
    \lbrace y\in \mathbb{R}^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\} \nonumber \\
    \min\{\varphi(y): y\in D\}=\min\{\varphi(y(x)): x\in [0,1]\} \nonumber
  \end{gather}
  \(y(x)\) is non-smooth function which continuously maps the segment \([0,1]\) to the hypercube \(D\).
  \begin{figure}[ht]
    \vspace*{-0.5cm}
    \subfloat{{\includegraphics[width=.35\textwidth]{peano3d.png} }}
    \subfloat{{\includegraphics[width=.35\textwidth]{peano2d.png} }}
  \end{figure}
\end{center}
\end{frame}

\begin{frame}
  \frametitle{Properties of the reduced problem}
  After applying the Peano-type evolvent \(\varphi(y(x))\) satisfies the uniform H{\"o}lder condition:
  \begin{displaymath}
    |\varphi(y(x_1))-\varphi(y(x_2))|\leq H{|x_1-x_2|}^{\frac{1}{N}}, x_1,x_2\in[0,1],
  \end{displaymath}
  \(\varphi(y(x))\) is non-smooth and has multiple local and \textbf{global} extremums even if \(\varphi(y)\) is unimodal.
  The latter problem is caused by loss of the information about \(N\)-d neighborhood after the transformation to the \(1\)-d space.

  \begin{figure}[ht]
    \begin{center}

    \vspace*{-0.5cm}
    \subfloat{\raisebox{.02\textheight}{\includegraphics[width=.3\textwidth]{parabaloid.png}}}
    \subfloat{\raisebox{.2\textheight}{\includegraphics[width=.05\textwidth]{arrow.jpg}}}
    \subfloat{{\includegraphics[width=.45\textwidth]{map_paraboloid.pdf}}}
  \end{center}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Basic optimization method}

  Optimization method generates search sequence \(\{x_k\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose \(p\) intervals \((x_{t_j-1}, x_{t_j})\) with the greatest characteristics and
    compute objective \(\varphi(y(x^{k+j}))\) in points chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1+j}=d(t)\in (x_{t_j-1}, x_{t_j}),\:j=\overline{1,p}
    \end{displaymath}
    \item[Step 4.] If \(x_{t_j}-x_{t_j-1}<\varepsilon\) for one of \(j=\overline{1,p}\), stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

\begin{frame}
  \frametitle{Test problems}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      Generator GKLS was employed to construct the sets of test problems:
      \begin{displaymath}
        \begin{matrix}
          f(x)=
          \left\{
          \begin{matrix}
          C_i(x), x \in S_i, i\in 2,\dots ,m \\
          \Vert x-T \Vert^2 + t, x\not\in S_2,\dots,S_m
          \end{matrix} \right.
        \end{matrix}
      \end{displaymath}

@@

\begin{frame}
  \frametitle{Cluster environment}
  \begin{center}

  The computational experiments have been carried out on the Lobachevsky supercomputer at  State University of Nizhni Novgorod.
  One node includes 2 Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB RAM.

  Each node runs under CentOS 7 Linux with GCC 4.8 compiler and Intel MPI library.
\end{center}

\end{frame}

\begin{frame}
  \frametitle{Evolvents comparison}
  %\begin{figure}[ht]
  %  \hspace*{-0.9cm}
  %  \subfloat[Minimal \(r\)]{{\includegraphics[width=.55\textwidth]{gklsS2d_opt_pt_op.pdf} }}
  %  \subfloat[\(r=5.0\)]{{\includegraphics[width=.55\textwidth]{gklsS2d_same_r_opt_pt_op.pdf} }} \hspace*{4cm}
  %  \caption{Operating characteristics on GKLS 2d Simple class}
  %\end{figure}
\end{frame}

\begin{frame}
  \frametitle{Evolvents comparison}
  %\begin{figure}[ht]
  %  \hspace*{-0.9cm}
  %  \subfloat[Minimal \(r\)]{{\includegraphics[width=.55\textwidth]{gklsS3d_opt_pt_op.pdf} }}
  %  \subfloat[\(r=4.5\)]{{\includegraphics[width=.55\textwidth]{gklsS3d_same_r_opt_pt_op.pdf} }}
  %  \caption{Operating characteristics on GKLS 3d Simple class}
  %\end{figure}
\end{frame}

\begin{frame}
  \frametitle{Choice of evolvent for the parallel algorithm}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}

\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}

\end{frame}

>>>>

\begin{frame}
  \frametitle{Methods}
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}
    \begin{itemize}
      \item Have complicated internal structure in multidimensional case;
      \item Usually store and use the whole history of trials accumulated during search;
      \item Require non-trivial parallelization schemes;
      \item Under some assumptions convergence to the global solution is guaranteed;
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}
    \begin{itemize}
      \item Have relative simple internal structure;
      \item Require constant amount of memory to store internal state of some random process or individuals of population;
      \item In most cases allows wide-scale parallelization by trials or by running many copies of the same method
      with different seeds in parallel.
      \item Convergence is guaranteed in probabilistic sense only.
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Methods}
  In this work considered the following solvers available in open-source:
  \begin{itemize}
    \item[$\square$] \textbf{Deteriminstic}
    \begin{itemize}
      \item DIRECT, DIRECT$l$;
      \item AGS, AGS$l$;
    \end{itemize}
    \item[$\square$] \textbf{Stochastic}
    \begin{itemize}
      \item Multi Level Single Linkage;
      \item StoGO;
      \item Differential Evolution;
      \item Controlled Random Search;
      \item Dual Simulated Annealing;
      \item Ant colony optimization;
    \end{itemize}
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Test problems}
  \begin{columns}
    \begin{column}{0.5\textwidth}
      Generator GKLS was employed to construct 8 sets of 100 test problems:
      \begin{displaymath}
        \begin{matrix}
          f(x)=
          \left\{
          \begin{matrix}
          C_i(x), x \in S_i, i\in 2,\dots ,m \\
          \Vert x-T \Vert^2 + t, x\not\in S_2,\dots,S_m
          \end{matrix} \right.
        \end{matrix}
      \end{displaymath}

@@

\begin{frame}
  \frametitle{Results of sequential methods}

  \begin{table}
  \begin{center}
  \caption{Averaged number of trials executed by sequential methods for solving the test
  optimization problems}
  \resizebox{\textwidth}{!}{%
    \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
      \hline
      & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
    \hline
    \(F_{GR}\)     & 193.1 &  \textbf{158.3} & 400.3 & 182.3 & 214.9 & 947.2 & 691.2 & 1257.3 &
  1336.8 \\
    \hline
    GKLS 2d Simple &  254.9 & 217.6 & 510.6 & \textbf{189.0} & 255.2 & 556.8 & 356.3 & 952.2
  & 1251.5 \\
    \hline
    GKLS 2d Hard   &  728.7 & \textbf{488.0} & 844.7 & 985.4 & 1126.7 & 1042.5 & 1637.9 &
  1041.1 & 2532.2 \\
    \hline
    GKLS 3d Simple &  1372.1 & 1195.3 & 4145.8 & \textbf{973.6} & 1477.8 & 4609.2 & 2706.5 &
  5956.94 & 3856.1 \\
    \hline
    GKLS 3d Hard   &  3636.1 & \textbf{1930.5} & 6787.0 & 2298.7 & 3553.3 & 5640.1 & 4708.4 &
  6914.3 & 7843.2 \\
    \hline
    GKLS 4d Simple &  26654.1 & 11095.7 & 37436.8 & \textbf{7824.3} & 15994.1 & 41514.3 &
  21417.9 & 19157.7 & 59895.4 \\
    \hline
    GKLS 4d Hard   &  54536.8 &  \textbf{23167.8} &  73779.3 &  23204.4 &  54489.9 &  80247.2
  &  68815.5 &  27466.1 &  109328.1  \\
    \hline
    GKLS 5d Simple &  29810.0 & 11529.0 & 143575.0 & \textbf{7166.5} & 13970.5 & 52647.6 &
  34255.3 & 73074.5 & 91580.4 \\
    \hline
    GKLS 5d Hard   &  113129.1 & 67652.7 & 165192.8 & \textbf{66327.4} & 164390.6 & 138766.2
  & 116973.1 & 105496.9 & 155123.8 \\
    \hline
  \end{tabular}}
    \label{tab:trials}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of sequential methods}
  \begin{table}
  \begin{center}
  \caption{Number of test optimization problems solved by sequential methods}
    \begin{tabular}{|l|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|{c}|}
      \hline
      & AGS & AGS\(l\) & CRS & DIRECT & DIRECT\(l\) & MLSL & SDA & DE & StoGO \\
    \hline
    \(F_{GR}\)     &  100 & 100 & 76  & 100 & 100 & 97  & 96  & 96  & 67\\
    \hline
    GKLS 2d Simple &  100 & 100 & 85  & 100 & 100 & 100 & 100 & 98  & 90\\
    \hline
    GKLS 2d Hard   &  100 & 100 & 74  & 100 & 100 & 100 & 93  & 85  & 77 \\
    \hline
    GKLS 3d Simple &  100 & 97  & 75  & 100 & 100 & 100 & 89  & 86  & 44 \\
    \hline
    GKLS 3d Hard   &  100  & 99   & 72   & 100  & 99   & 100  & 88   & 77   & 43 \\
    \hline
    GKLS 4d Simple &  100 & 100 & 46  & 100 & 100 & 94  & 78  & 59  & 16 \\
    \hline
    GKLS 4d Hard   &  100 & 100 & 47  & 99  & 97  & 94  & 72  & 32  & 10  \\
    \hline
    GKLS 5d Simple &  100 & 100 & 68  & 100 & 100 & 98  & 100 & 77  & 9  \\
    \hline
    GKLS 5d Hard   &  97  & 99  & 42  & 100 & 90  & 79  & 84  & 48  & 8 \\
    \hline
    \end{tabular}
    \label{tab:solved}
  \end{center}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Basic optimization method}
  Optimization method generates search sequence \(\{x_k\}\) and consists of the following steps:
  \begin{enumerate}
    \setlength{\itemindent}{.1in}
    \item[Step 1.] Sort the search information (one-dimensional points) in increasing order.
    \item[Step 2.] For each interval \((x_{i-1}, x_i)\) compute quantity \(R(i)\), called characteristic.
    \item[Step 3.] Choose \(p\) intervals \((x_{t_j-1}, x_{t_j})\) with the greatest characteristics and
    compute objective \(\varphi(y(x^{k+j}))\) in points chosen using the decision rule \(d\):
    \begin{displaymath}
      x^{k+1+j}=d(t)\in (x_{t_j-1}, x_{t_j}),\:j=\overline{1,p}
    \end{displaymath}
    \item[Step 4.] If \(x_{t_j}-x_{t_j-1}<\varepsilon\) for one of \(j=\overline{1,p}\), stop the method.
  \end{enumerate}
  \textit{\footnotesize	{Detailed description: Strongin R.G., Sergeyev Ya.D.: Global optimization with non-convex constraints. Sequential and parallel algorithms (2000), Chapter 7}}
\end{frame}

\begin{frame}
  \begin{center}
  \frametitle{Dimension reduction}
  Peano-type curve \(y(x)\) allows to reduce the dimension of the original problem:
  \begin{gather}
    \lbrace y\in \mathbb{R}^N:-2^{-1}\leqslant y_i\leqslant 2^{-1},1\leqslant i\leqslant N\rbrace=\{y(x):0\leqslant x\leqslant 1\} \nonumber \\
    \min\{\varphi(y): y\in D\}=\min\{\varphi(y(x)): x\in [0,1]\} \nonumber
  \end{gather}
  \(y(x)\) is non-smooth function which continuously maps the segment \([0,1]\) to the hypercube \(D\).
  \begin{figure}[ht]
    \vspace*{-0.5cm}
    \subfloat{{\includegraphics[width=.35\textwidth]{peano3d.png} }}
    \subfloat{{\includegraphics[width=.35\textwidth]{peano2d.png} }}
  \end{figure}
\end{center}
\end{frame}

\begin{frame}
  \frametitle{Parallel Globalizer solver}
  \begin{figure}[ht]
    \includegraphics[width=0.95\textwidth]{evolvents_parallel.pdf}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Parallel MIDACO solver}
\end{frame}

\begin{frame}
  \frametitle{Cluster environment}
  \begin{center}
  The computational experiments have been carried out on the Lobachevsky supercomputer at  State University of Nizhni Novgorod.
  One node includes 2 Intel Sandy Bridge E5-2660 2.2 GHz processors and 64 GB RAM.

  Each node runs under CentOS 7 Linux with GCC 4.8 compiler and Intel MPI library.
\end{center}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{table}
    \centering
    \caption{Averaged numbers of iterations executed by the parallel algorithms for solving the test
  optimization problems}
    \label{tab:iterations}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
      I & \textbf{1 cluster node}
        & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
      &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
      \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
  &   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
      \hline \noalign{\smallskip}
  III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
  & & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
      \hline \noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
  & & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
      \hline \noalign{\smallskip}
  V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
  & & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
      \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{table}
    \centering
    \caption{Speedup of parallel computations executed by the parallel algorithms}
    \label{tab:speedup}
    \begin{tabular}{cccccccc}
      \cline{3-8}\noalign{\smallskip}
      \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
  \multicolumn{2}{c}{MIDACO}   \\
      \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
      \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
  \textit{Hard}  \\
      \noalign{\smallskip}\hline
  I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
      & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
    &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
    \hline \noalign{\smallskip}
  II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
    & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
    &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
      \noalign{\smallskip}\hline	\noalign{\smallskip}
  III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
  &   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
  VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
  & & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
      \noalign{\smallskip}\hline
      V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
      & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
      & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
          \noalign{\smallskip}\hline
    \end{tabular}
  \end{table}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_1.pdf}}\label{fig:speedup_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{speedup_midaco_globalizer_p_16.pdf}}\label{fig:speedup_p16}}
    \caption{Speedup demonstrated by the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

\begin{frame}
  \frametitle{Results of applying the parallel algorithm}
  \begin{figure}[ht]
    \centering
    \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_1.pdf}}\label{fig:time_p1}}
    \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{time_midaco_globalizer_p_16.pdf}}\label{fig:time_p16}}
    \caption{Averaged execution time of the parallel algorithms when solving problems from the GKLS 4d Hard class}
  \end{figure}
\end{frame}

%\begin{frame}
%  \frametitle{Results of applying the parallel algorithm}
%\end{frame}
